Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-02 doc/paper-lcr.tex
started theory section

% Packages:
\input psfig.sty
\usepackage{xspace}
\usepackage{graphicx}

@@

\author[Collett \etal]{%
  Thomas~J.~Collett$^{1}$\thanks{\collettemail},
  Philip~J.~Marshall$^{2}$,
  Matthew~W.~Auger$^{1}$,
  Stefan~Hilbert$^{3}$,
\newauthor{%
  Sherry~H.~Suyu$^{4}$,
  Zachary~Greene$^{4}$,
  Tommaso~Treu$^{4}$\thanks{\packard},
  Christopher~D.~Fassnacht$^{5}$,}
\newauthor{%
  L\`eon~V.~E.~Koopmans$^{6}$,
  Roger~D.~Blandford$^{3}$} 
  \medskip\\
  $^1$\ioa\\
  $^2$\oxford\\
  $^3$\kipac\\
  $^4$\ucsb\\
  $^5$\davis\\
  $^6$\kapteyn
}

@@

Strong gravitational lenses can provide high precision cosmological distance
measurements; one of the most important sources of systematic error in these
estimates is the additional mass along the line of sight to the source. Weak
lensing by galaxies and groups in this beam can contribute an ``external
convergence'' $\xkappa$ that dominates the  uncertainty in the inferred
distances; likewise, it may also contaminate estimates of the luminosity of
objects at high redshift (such as reionisation-era proto-galaxies, and type Ia
supernovae standard candles).  We characterise this uncertainty by marginalising
over a probability density function for $\xkappa$: here, we investigate the use
of a simple halo model to estimate $\Pr(\xkappa)$ given noisy estimates of the
photometric redshift and stellar mass of galaxies observed in optical imaging
surveys of the field in question. Tests using mock catalogs from the Millenium
Simulation and comparing to ray-traced $kappa$ values, we find that \ldots 
The $\Pr(\xkappa)$ from this reconstruction process are narrower than the
expected distribution of $\xkappa$ values over an ensemble of time delay lenses:
incorporating the photometric catalogs in a joint analysis of 100 lenses 
leads to an increase in precision of factor of 1.X over simple averaging. We
investigate several sources of bias and determine \ldots
\end{abstract}

@@

Time delay distances, and the mass sheet degeneracy - how kappa is degenerate in
strong lens models (and magnifications). Weak lensing by NFW halos. Accounting
for mass along the line of sight.  Approximations to full ray tracing: combining
kappas: simple sum, Keeton scaling, comments. Example ray traced kappa map from Stefan.

>>>>

% Packages:
\input psfig.sty
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}

@@

\author[Collett \etal]{%
  Thomas~E.~Collett$^{1}$\thanks{\collettemail},
  Philip~J.~Marshall$^{2}$,
  Matthew~W.~Auger$^{1}$,
  Stefan~Hilbert$^{3}$,
\newauthor{%
  Sherry~H.~Suyu$^{4}$,
  Zachary~Greene$^{4}$,
  Tommaso~Treu$^{4}$\thanks{\packard},
  Christopher~D.~Fassnacht$^{5}$,}
\newauthor{%
  L\`eon~V.~E.~Koopmans$^{6}$,
  Roger~D.~Blandford$^{3}$} 
  \medskip\\
  $^1$\ioa\\
  $^2$\oxford\\
  $^3$\kipac\\
  $^4$\ucsb\\
  $^5$\davis\\
  $^6$\kapteyn
}

@@

Strong gravitational lenses can provide high precision cosmological distance
measurements; one of the most important sources of systematic error in these
estimates is the additional mass along the line of sight to the source. Weak
lensing by galaxies and groups in this beam can contribute an ``external
convergence'' $\xkappa$ that dominates the  uncertainty in the inferred
distances; likewise, it may also contaminate estimates of the luminosity of
objects at high redshift (such as reionisation-era proto-galaxies, and type Ia
supernovae standard candles).  We characterise this uncertainty by marginalising
over a probability density function for $\xkappa$: here, we investigate the use
of a simple halo model to estimate $\Pr(\xkappa)$ given noisy estimates of the
photometric redshift and stellar mass of galaxies observed in optical imaging
surveys of the field in question. Tests using mock catalogs from the Millenium
Simulation and comparing to ray-traced $\kappa$ values, we find that \ldots 
The $\Pr(\xkappa)$ from this reconstruction process are narrower than the
expected distribution of $\xkappa$ values over an ensemble of time delay lenses:
incorporating the photometric catalogs in a joint analysis of 100 lenses 
leads to an increase in precision of factor of 1.X over simple averaging. We
investigate several sources of bias and determine \ldots
\end{abstract}

@@

%Time delay distances[✓], and the mass sheet degeneracy - how kappa is degenerate in
%strong lens models (and magnifications). Weak lensing by NFW halos. Accounting
%for mass along the line of sight.  Approximations to full ray tracing: combining
%kappas: simple sum, Keeton scaling, comments. Example ray traced kappa map from Stefan.

Strong lensing occurs when a massive object and background source are almost perfectly
aligned along a line of sight. The light from the background source is deflected by the
lens galaxy; this deflection allows multiple images of the background source to form
at stationary points of the time delay function. For a strong lens system in an otherwise
homogeneous Robertson--Walker universe, the excess time delay of an image at angular position
$\bmath{\theta}=(\theta_1,\theta_2)$ with unlensed source position $\bmath{\beta}=(\beta_1,\beta_2)$ is
\be \label{eq:T} 
\Delta t(\bmath{\theta},\bmath{\beta}) = \frac {1}{c} \frac{D_{\rm d} D_{\rm s}}{D_{\rm ds}} (1+z_{\rm d})\, \phi(\bmath{\theta},\bmath{\beta}),
\ee
where $z_{\rm d}$ is the redshift of the lens, $\phi(\bmath{\theta},\bmath{\beta})$ is
the Fermat potential, and $D_{\rm d}$, $D_{\rm s}$, and $D_{\rm ds}$
are, respectively, the angular diameter distance from observer to the lens, from observer
to the source, and from the lens to the source. The Fermat potential is
defined as
\be \label{eq:FP}
\phi(\bmath{\theta},\bmath{\beta})\equiv \left[\frac{(\bmath{\theta}-\bmath{\beta})^2}{2}-\psi(\bmath{\theta}) \right]. 
\ee
The first term in Eq. \ref{eq:FP} comes from the geometric path difference as a result of
the deflection, and the second term is the gravitational delay
described by the lens potential $\psi(\bmath{\theta})$.

The lens potential, $\psi(\bmath{\theta})$, is related to the projected dimensionless
surface mass density, $\kappa(\bmath{\theta})$, and the physical projected surface mass density,
$\Sigma(D_{\rm d} \bmath{\theta})$, by 
\be \label{eq:psiKappaDiffRelan}
\kappa(\bmath{\theta})= \frac {\Sigma(D_{\rm d} \bmath{\theta})} {\Sigma_{\rm cr}}=\frac{1}{2}\nabla^2\psi(\bmath{\theta}),
\ee
where 
\be \label{eq:sigcrit} 
\Sigma_{\rm cr} \equiv \frac{c^2 D_{\rm s}}{4 \pi G D_{\rm d} D_{\rm ds}}.
\ee

Equation \ref{eq:T} can be rephrased as
\be
\Delta t(\bmath{\theta},\bmath{\beta})  =  \frac{\Dt}{c},
\ee 
where $\Dt \equiv (1+\zd) D_{\rm d} D_{\rm s}/D_{\rm ds}$ is the time-delay distance.
Angular diameter distances are given by
\begin{equation}
D_{\mathrm{ij}}~=~ {c/H_0 \over {(1+z_{\mathrm{j}})}}\left( {\mathrm{sinn}\!\left( \sqrt{|\Omega_k|} \int_{z_{\mathrm{i}}}^{z_{\mathrm{j}}} {\mathrm{d} z \over E(z)}\right)\over \sqrt{|\Omega_k|}}\right)
\label{eq:Da}
\end{equation}
where $\mathrm{sinn}(x)~=~\sin(x)$, $x$, or $\sinh(x)$ for open, flat, or closed universes respectively, and $E(z)$ is the normalised Hubble parameter:
\begin{subequations}\label{Eq:Ez}
\begin{align}
E(z) &\equiv {H(z) \over H_0}\\
E^{\Lambda{CDM}}&~=~\!\sqrt{{\Omega_{\mathrm{M}}(1\!+\!z)^{3}\!+\!\Omega_k(1\!+\!z)^2\!+\!\Omega_{\Lambda}}}
\end{align}
\end{subequations}
Since angular diameter distances are proportional to the inverse of the Hubble constant, $1/H_0$,
so is $\Dt$ proportional to $1/H_0$, hence a measurement of $\Dt$ can be used to infer $H_0$
and also the other parameters in Equations \ref{eq:Da} and \ref{eq:Ez}.

\subsection{Mass-sheet degeneracy}
\label{sec:H0theory:MSD}

A dificulty with using time delay distances to contrain $H_0$ is the so called
Mass-sheet degeneracy \citep[see e.g.][for details]{Falco1985,Schneider2006}.
This is a degeneracy in the strong lens modeling corresponding to the addition of a
mass sheet that contributes a convergence and zero shear which combined with a
re-scaling of the original mass distribution leaves the predicted image positions and
magnifications unchanged. Given a lens
model $\kappa_{\rm lens}(\vec{\theta})$, an alternative model described by the
transformation $\tilde{\kappa}_{\rm lens}(\vec{\theta}) = \lambda + (1-\lambda)
\kappa_{\rm lens}(\vec{\theta})$, where $\lambda$ is a constant,
would fit the image positions, shapes and relative fluxes equally well. Physically $\lambda$
is the convergence of the mass-sheet and can take any real value.

Since strong lenses sit in our real Universe, they are not truely isolated;
there is some mass along the line of sight coming from intervening structures.
These structures may be physically associated with the lens e.g. other members
of a group or cluster, or they may be the result of chance alignments at any redshift. 
Whilst it is rare for three galaxies to line up well enough for both of the background sources
to be strongly lensed \citep{gavazzi2008,collett2012}, the shear scale of dark matter halos
makes a partial allignment -- with the additional structure acting as a weak lens -- 
a near certainty, indeed \citet{valewhite2003} and \citet{hilbert2007} find that there
are no empty lightcones in a realistic universe, as such the

it is clear that 
the mass-sheet degeneracy corresponds to a degeneracy between this
external convergence ($\kext$) and the mass normalization of the lens 
galaxy.\footnote{To be specific, the prescription that we adopt for
  combining the effects of many mass sheets at redshifts $z_i$ with
  surface mass densities $\Sigma_i$ is $\kext=\frac{4\pi
    G}{c^2}\displaystyle\sum_i\frac{\Sigma_i(D_i\vec\theta)D_i D_{i{\rm s}}}{D_{\rm s}}$.}

Despite the invariance of the image positions, shapes and relative fluxes
under a mass-sheet transformation, the relative Fermat potential between the
images changes according to $\Delta \phi_{\rm trans}(\vec{\theta},
\vec{\beta}_{\rm trans}) = (1-\lambda) \Delta \phi_{\rm
  model}(\vec{\theta},\vec{\beta}_{\rm model})$.  Therefore, given measured
relative time delays $\Delta t$, which are inversely proportional to $H_0$ and
proportional to the relative Fermat potential (Equation \ref{eq:TsimpH0}), the
transformed model $\kappa_{\rm trans}$ would lead to an $H_0$ that is a factor
$(1-\lambda)$ lower than that of the initial $\kappa_{\rm model}$ (for fixed
$\OM$, $\OL$, and $w$).  In other words, if there is physically any external
convergence $\kext$ due to the lens' local environment or mass structure along the
line of sight
to the lens system that is not incorporated in the lens
modeling, then
\be 
\label{eq:MassSheet:H0bias}
H_0^{\rm{true}}=(1-\kext) H_0^{\rm{model}}.
\ee

This degeneracy is present because lensing observations only deliver
relative positions and fluxes.  The degeneracy can be broken, allowing us to
measure $H_0$, if (i) we
know the magnitude or angular size of the source in absence of lensing,
(ii) we have information on
the mass normalization of the lens,
or (iii) we can compare the measured shear in the lens with the
observed distribution of mass to calibrate $\kext$.
For most of the strong lens systems including \ourlens, 
case (i) does not apply, so
circumventing the mass-sheet degeneracy requires the input of more
information, either about the lensing galaxy itself, or its three-dimensional
environment.
We distinguish between two kinds of mass sheets: \textit{internal}
and \textit{external}.  
%
Internal mass 
sheets, which are physically associated with the lens galaxy, are due
to nearby, physically associated galaxies, 
groups or clusters which, crucially, 
affect the stellar dynamics of the
lens galaxy. External mass sheets describe mass
distributions that are not physically associated with the
lens galaxy and, by definition, do not affect the stellar dynamics.  
Typically these will 
lie along the line of sight to the lens \citep{FassnachtEtal06}. 
We identify $\kext$ as the net convergence of this external mass sheet.

Two methods for breaking the mass-sheet degeneracy are then: 
\renewcommand{\theenumi}{\roman{enumi}}
\begin{enumerate}
\item \textit{Stellar dynamics of the lens galaxy.}  Stellar dynamics
  can be used jointly with lensing to break the internal mass-sheet degeneracy
  by providing an estimate of the enclosed mass at a radius different
  from the Einstein radius, which is approximately the radius of the
  lensed images from the lens galaxy
  \citep[e.g.,][]{GroginNarayan96a, GroginNarayan96b, TonryFranx99,
    KoopmansTreu02, TreuKoopmans02, BarnabeKoopmans07}.  We note that
  for a given stellar velocity dispersion, there is a degeneracy in
  the mass and the stellar orbit anisotropy (which characterizes the amount of
  tangential velocity dispersion relative to radial dispersion).  Nonetheless,
  the mass-isotropy degeneracy is nearly orthogonal to the mass-sheet
  degeneracy, so a combination of the mass within the effective radius
  (from the stellar velocity dispersion) and the mass within the Einstein
  radius (from lensing) effectively breaks both the mass-isotropy and
  the internal mass-sheet degeneracies. We describe how this works within the
  context of our chosen mass model in Section~\ref{sec:H0theory:Dynamics}
  below.
\item \textit{Studying the environment and the line of sight to the lens galaxy.}  
  Observations of the field around lens galaxies allow a rough picture of the
  projected mass distribution to be built up. Many lens galaxies lie in galaxy
  groups, which can be identified either by their spectra or, more cheaply
  (but less accurately), by their colors and magnitudes.    By modeling the
  mass distribution of the groups and galaxies in the lens plane and along the
  line of sight  to the lens galaxy, one can estimate the external
  convergence~$\kext$  at the redshift of the lens \citep[e.g.][and references
  therein]{MomchevaEtal06, FassnachtEtal06, AugerEtal07}.  The group modeling
  requires (i) identification of the  galaxies that belong to the group of the
  lens galaxy, and (ii) estimates of the group centroid and velocity
  dispersion.  A number of recipes can be followed. For example,
  \citet{KeetonZabludoff04} considered two extremes: (i) the group is
  described by a single smooth mass distribution, and (ii) the masses are
  associated with individual galaxy group members with no common halo.  The
  realistic mass distribution for a galaxy group should be somewhere between
  these two extremes.  
  The experience to date  is that 
  modeling lens environments accurately is very difficult, with uncertainties
  of 100\% typical \citep[e.g.][]{MomchevaEtal06, FassnachtEtal06}.    
  In Section \ref{sec:LensEnv}, we describe an
  alternative approach for quantifying the external convergence in a
  statistical manner: ray-tracing through numerical simulations of large-scale
  structure \citep{HilbertEtal07}.
  In this section 
  we also present a first attempt at tailoring the ray-tracing results 
  to our one line of sight, using the relative galaxy number counts in the
  field.
\end{enumerate}

We emphasize that the mass-sheet degeneracy is simply one of the
several parameter degeneracies in the lens modeling that has been
given a special name.  When power-laws ($\kappa \sim b R^{1-\slope}$,
where $R$ is the radial distance from the lens center, $b$ is the
normalization of the lens, and $\slope$ is the radial slope in the
mass profile) are used to describe the lens mass distribution, one
often finds a $H_0$-$\slope$ degeneracy in addition to the
$H_0$-$b$-$\kext$ (mass-sheet) degeneracy (for fixed $\OM$, $\OL$ and
$w$; more generally, $\tdist$ would be in place of $H_0$).  These two
degeneracies are of course related via $H_0$.  The $H_0$-$\slope$
degeneracy primarily occurs in lens systems with symmetric
configurations due to a lack of information on $\slope$.  In contrast,
lens systems with images spanning a range of radii or with extended
images provide information on $\slope$
\citep[e.g.][]{WucknitzEtal04,DyeEtal08}, and so the $H_0$-$\slope$
degeneracy is broken.  Nonetheless, the $H_0$-$b$-$\kext$ degeneracy
is still present unless we provide information from dynamics and lens
environment studies.
